---
title: "Data cleaning, tidy data"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---


## Clean data:
- one data, one variable
- one row per observation
- data and type matches
- same name for same categories
- clean factors, if necessary

Exercise

1. Clean up the following dataset, using base R:
messy_data.xlsx

```{r}

```



## Tidy data

by Hadley Wickham
- long over wide format
- data is in tibbles
- tidyverse offers various functions to easily manipulate the data
- new concept: piping.

Examples:


```{r}
library(dplyr)
library(tidyr)
library(tibble)
data(mtcars)

mtcars <- as_tibble(rownames_to_column(mtcars))

filtered <-mtcars %>%
  filter(gear<5) 

filtered %>% 
  group_by(carb) %>%
  summarise(qsec_mean=mean(qsec))

#it is possible to combine as many steps as needed without saving it in a separate object

mtcars %>%
  filter(gear<5) %>%
  group_by(carb) %>%
  summarise_all(list(sum=sum, mean=mean))

mtcars %>%
  mutate(cyl_2=if_else(cyl==8, 7, cyl))


mtcars %>%
  mutate(gear=gear-1)

mtcars %>%
  pivot_longer(-cars, names_to = "name", values_to = "values")

```


2. Clean up the following dataset, using as much tidyverse as possible:
messy_data.xlsx

```{r}

```


## Merging data

`merge()` funtction from base R
`join()` function from tidyvers

```{r}

band_members %>% full_join(band_instruments)

merge(band_members, band_instruments, by="name", all=T, sort=F)

```

