---
title: "Data cleaning, tidy data"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---


## Clean data:
- one data, one variable
- one row per observation
- data and type matches
- same name for same categories
- clean factors, if necessary

Exercise

1. Clean up the following dataset, using base R:
messy_data.xlsx
messy_data.txt


```{r}
messy_data <- read.delim("~/GitHub/R_course_basic/data/messy_data.txt")

messy_data$Patient.id <- toupper(messy_data$Patient.id)
messy_data$Patient.id <- gsub(" ", "_", messy_data$Patient.id)
messy_data <- messy_data[1:18,]

messy_data$X <- NULL

Sys.setlocale("LC_TIME", "English")
#help: https://www.stat.berkeley.edu/~s133/dates.html
messy_data$Date.of.surgery <- as.Date(messy_data$Date.of.surgery, format="%d-%b-%y")
messy_data$Date.of.birth <- as.Date(as.Date( messy_data$Date.of.birth, format="%d/%m/%Y"))
messy_data$Date.of.diagnosis <- as.Date(messy_data$Date.of.diagnosis, format="%d/%m/%Y")
messy_data$Time.of.death <- as.Date(messy_data$Time.of.death, format="%d-%b-%y")



```



## Tidy data

by Hadley Wickham
- long over wide format
- data is in tibbles
- tidyverse offers various functions to easily manipulate the data
- new concept: piping.

Examples:


```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(magrittr)
data(mtcars)

mtcars <- as_tibble(rownames_to_column(mtcars))

filtered <- mtcars %>%
  filter(gear<5) 

filtered %>% 
  group_by(carb) %>%
  summarise(qsec_mean=mean(qsec))

#it is possible to combine as many steps as needed without saving it in a separate object

mtcars %>%
  filter(gear<5) %>%
  group_by(carb) %>%
  summarise(sum_drat=sum(drat), mean_drat=mean(drat))

mtcars %>%
  mutate(cyl_2=if_else(cyl==8, 7, cyl))


mtcars %>%
  mutate(gear=gear-1)

mtcars %>%
  pivot_longer(-rowname, names_to = "name", values_to = "values")

```


2. Clean up the following dataset, using as much tidyverse as possible:
messy_data.xlsx

```{r}

messy_data <- messy_data %>% 
  select_all(~gsub(" ", ".", .))

messy_data <- messy_data %>%
  mutate(Patient.id=toupper(Patient.id)) %>%
  mutate(Patient.id=gsub(" ", "_", Patient.id), 
         Age=round(as.numeric((Date.of.diagnosis-Date.of.birth)/365), digits = 0),
         N=substr(TNM, 3, 4), 
         M=substr(TNM, 5, 6), 
         TNM=NULL)
         #enzyme.level=if_else(enzyme.level>600, NA, enzyme.level))


messy_data$enzyme.level[messy_data$enzyme.level>600] <- NA


```


## Merging data

`merge()` funtction from base R
`join()` function from tidyvers

```{r}

band_members %>% full_join(band_instruments)

merge(band_members, band_instruments, by="name", all=T, sort=F)

```

